---
title: "APIS Canopy Calculation"
author: "Ryan Pienaar"
date: "`r Sys.Date()`"
output: html_document
---

```{r}

library(dplyr)
library(readr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(stringr)
library(sjPlot)
library(sf)
library(tidyverse)

```

# Pauli lab ground sampling

```{r}
# Load your CSV
df <- read.csv("Trees_Sites.csv")  # Replace with your actual file path
canopy <- read_csv("CanopyHeightMonica.csv")

```

Correct species names
```{r}

head(df)
unique(df$Species)


df <- df %>%
  mutate(Species= recode(Species,
                          "Aspen sp." = "Aspen",
                         "BTA" = "Aspen",
                         "ASPEN SP." = "Aspen",
                          "Maple sp." = "Maple",
                         "Maple SP." = "Maple",
                         "red maple" = "Maple",
                         "RMN" = "Maple",
                         "RNM" = "Maple",
                         "REM" = "Maple",
                         "RM" = "Maple",
                          "SM" = "Maple",
                         "SRM" = "Maple",
                         "sugar maple"= "Maple",
                         "MAPLE SP." = "Maple",
                         "maple? "  = "Maple",
                          "QA" = "Aspen",
                         "aspen"  = "Aspen",
                          "BAL " = "Balsam",
                         "VAL " = "Balsam",
                         "VAL" = "Balsam",
                         "balsam" = "Balsam",
                         "BALSM" = "Balsam",
                         "BIRCH SP." = "Birch",
                         "YB." = "Birch",
                         "YB" = "Birch",
                         "yellow birch" = "Birch",
                         "PB" = "Birch",
                         "paper birch" = "Birch",
                         "BAL" = "Balsam",
                         "C "= "Cedar",
                         "C"= "Cedar",
                         "cedar"= "Cedar",
                         "H" ="Hemlock",
                         "H " = "Hemlock",
                         "WP" = "Pine",
                         "white pine?" = "Pine",
                         "IW" = "Ironwood",
                         "BASS" = "Basswood",
                         "WO?" = "Oak",
                         "WO" = "Oak",
                         "RO" = "Oak",
                         "O" = "Oak",
                         "BS" = "Spruce",
                         "SPRUCE SP." = "Spruce",
                         "RP" = "Pine",
                         "T" = "Tamarack",
                         "beech? Picture" = "Beech",
                         "BA" = "Ash",
                         "BA " = "Ash",
                         "BL" = "Black locust",
                         "ELM" = "Elm",
                         "UNKNWN" = "Unknown",
                         "UNKWN" = "Unknown",
                         "UNKNWN, APPLE?" = "Unknown"))
                                      
```


```{r}
# Convert Height and DBH to numeric (in case they're read as characters)
df <- df %>%
  mutate(
    Height = as.numeric(Height),
    DBH = as.numeric(DBH)
  )

df <- df %>%
  filter(tolower(Vigor) != "snag")

df <- df %>%
  mutate(Site = str_replace(Site, "^Oak", "OAK"),
         Site = str_replace(Site, "^Sand", "SAND"))

# Filter to only trees within 1 SD of max height (likely canopy trees)
# canopy_trees <- df %>%
#   group_by(Site) %>%
#   mutate(
#     max_height = max(Height, na.rm = TRUE),
#     sd_height = sd(Height, na.rm = TRUE)
#   ) %>%
#   filter(Height >= (max_height - sd_height)) %>%
#   ungroup()


head(df)
head(canopy)
colnames(canopy)[3] <- "Site"

df <- df |>  left_join(canopy %>% select(Site, b1), by = "Site") %>%
  rename(CanopyHeight = b1)  # Rename b1 to Height

canopy_trees <- df |> filter(Height >= CanopyHeight)

```



```{r}
# Compute canopy area proxy
canopy_trees <- canopy_trees %>%
  mutate(canopy_area = (DBH^2))  

# Summarize by species at each site
species_canopy <- canopy_trees %>%
  group_by(Site, Species) %>%
  summarise(species_area = sum(canopy_area, na.rm = TRUE), .groups = "drop")

# Total canopy area per site
total_canopy <- species_canopy %>%
  group_by(Site) %>%
  summarise(total_area = sum(species_area), .groups = "drop")


#Add site coords back
site_coords <- df %>%
  select(Site, Lat, Long) %>%
  distinct()

canopy_proportion <- species_canopy %>%
  left_join(total_canopy, by = "Site") %>%
  mutate(proportion = species_area / total_area) %>%
  left_join(site_coords, by = "Site")  # <-- this keeps the lat/long


### Dominant species stands
stands <- filter(canopy_proportion, proportion >= 0.5)
write.csv(stands, "Stands.csv")

```


I want to determine the DBH cut off at which the tree is likely to be in the canopy

```{r}

ggplot(data= df, aes(x= DBH, y =Height, colour = Species))+
  geom_point()


# Fit the mixed-effects model
mod <- glm(Height ~ DBH * Species, data = df)

# Inspect the model
tab_model(mod)

# Define canopy height threshold
cutoff <- mean(canopy_trees$Height, na.rm = TRUE) + sd(canopy_trees$Height, na.rm = TRUE)



```



#NPS Ground Sampling

Load NPS data
```{r}

NPS <- read_csv("Y:/rpienaar/Apostle Islands/InputData/APIS_Veg_2021.csv")
NPSCanopy <- read.csv("CanopyHeightNPS.csv")

colnames(NPSCanopy)[2] <- "Plot"

NPS <- NPS |> left_join(NPSCanopy |> select(Plot, b1), by = "Plot") |>
   rename(CanopyHeight = b1)

NPS <- NPS |> mutate(genus= stringr::word(Accepted_LatinName, 1))

genus_common <- tibble(
  genus = c("Acer", "Abies", "Betula", "Thuja", "Tsuga",
            "Quercus", "Fraxinus", "Populus", "Ostrya", "Picea", "Tilia", "Ulmus",
            "Prunus", "Sorbus", "Pinus", "unknown"),
  common_name = c("Maple", "Balsam", "Birch", "Cedar", "Hemlock",
                  "Oak", "Ash", "Aspen", "Ironwood", "Spruce", "Basswood", "Elm",
                  "Unknown", "Unknown", "Pine", "Unknown")
)



NPS <- NPS %>%
  left_join(genus_common, by = "genus")


NPS <- NPS |> select(Accepted_LatinName, Diameter_cm, Status, CanopyHeight, Plot, genus, common_name, Northing, Easting )

```
Calculate heigth based on model from Monica

```{r}

#rename columns to match input data
colnames(NPS)[2] <- "DBH"
colnames(NPS)[7] <- "Species"

NPS$DBH <- as.numeric(NPS$DBH)
NPS$Species <- as.factor(NPS$Species)

NPS$Height <- predict(mod, newdata = NPS)


```


```{r}

NPS$Plot <- as.factor(NPS$Plot)
NPS <- NPS |> filter(Status == "Live")

NPScc <- NPS |> filter(Height >= CanopyHeight)

NPScc$canopy <- NPScc$DBH^2

# Summarize by species at each site
NPSspecies_canopy <- NPScc %>%
  group_by(Plot, Species) %>%
  summarise(species_area = sum(canopy, na.rm = TRUE), .groups = "drop")

# Total canopy area per site
NPStotal_canopy <- NPSspecies_canopy %>%
  group_by(Plot) %>%
  summarise(total_area = sum(species_area), .groups = "drop")

#Add site coords back
NPSsite_coords <- NPS %>%
  select(Plot, Northing, Easting) %>%
  distinct()

NPScanopy_proportion <- NPSspecies_canopy %>%
  left_join(NPStotal_canopy, by = "Plot") %>%
  mutate(proportion = species_area / total_area) %>%
  left_join(NPSsite_coords, by = "Plot")  # <-- this keeps the lat/long

NPScanopy_proportion$Northing <- as.numeric(NPScanopy_proportion$Northing)
```





Add common name

```{r}

NPScanopy_proportion |> filter(proportion > 0.5)

NPScanopy_proportion |> filter(Species == "Oak")


```


Now I have the 2 files of the 2 sources with the proportions of canopy cover, I should put them together to make one file

```{r}



head(canopy_proportion)

head(NPScanopy_proportion)

crs_utm <- 32615  
# Convert UTM to lat long
dfutmsf <- st_as_sf(NPScanopy_proportion, coords = c("Northing", "Easting"), crs = crs_utm)

dflatlong <- st_transform(dfutmsf, crs = 4326)

# Extract lat/long into columns
NPSprop <- dflatlong %>%
  mutate(Lat = st_coordinates(.)[,2],
          Long = st_coordinates(.)[,1]) %>%
  st_drop_geometry()


# Combine dfs
head(canopy_proportion)

head(NPSprop)

#rename cols
colnames(NPSprop)[1] <- "Site"
Combo <- rbind(canopy_proportion, NPSprop)

```


```{r}

ComboDom <- Combo |> filter(proportion >= 0.5) 

Combo <- Combo |> group_by(Site, total_area, Lat, Long, Species) %>%
  summarise(prop_area = sum(species_area) / unique(total_area), .groups = "drop") %>%
  pivot_wider(
    names_from = Species,
    values_from = prop_area)
    
Combo[is.na(Combo)] <- 0

head(Combo)


write.csv(ComboDom, "MandNPSdata.csv")
write.csv(Combo, "SpeciesProps.csv")

```

Summary of data I have

```{r}
library(gt)

Combo |> pivot_longer(cols = Aspen:Beech, names_to = "Species", values_to = "Cover") %>%
  mutate(
    Category = case_when(
      Cover == 1 ~ "Pure stands",
      Cover > 0 & Cover < 1 ~ "Mixed stands",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Category != "Other") %>%
  count(Species, Category) %>%
  pivot_wider(names_from = Category, values_from = n, values_fill = 0) |>
  gt()

```


Finish

```{r}


```

