---
title: "APIS Canopy Calculation"
author: "Ryan Pienaar"
date: "`r Sys.Date()`"
output: html_document
---

```{r}

library(dplyr)
library(readr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(stringr)
#library(sjPlot)
library(sf)
library(tidyverse)

```

# Pauli lab ground sampling

```{r}
# Load your CSV
df <- read.csv("Trees_Sites.csv")  # Replace with your actual file path
#canopy <- read_csv("CanopyHeightMonica.csv") # This is an extraction from GEDI canopy height layer

```

Correct species names
```{r}

head(df)
unique(df$Species)


df <- df %>%
  mutate(Species= recode(Species,
                          "Aspen sp." = "Aspen",
                         "BTA" = "Aspen",
                         "ASPEN SP." = "Aspen",
                          "Maple sp." = "Maple",
                         "Maple SP." = "Maple",
                         "red maple" = "Maple",
                         "RMN" = "Maple",
                         "RNM" = "Maple",
                         "REM" = "Maple",
                         "RM" = "Maple",
                          "SM" = "Maple",
                         "SRM" = "Maple",
                         "sugar maple"= "Maple",
                         "MAPLE SP." = "Maple",
                         "maple? "  = "Maple",
                          "QA" = "Aspen",
                         "aspen"  = "Aspen",
                          "BAL " = "Balsam",
                         "VAL " = "Balsam",
                         "VAL" = "Balsam",
                         "balsam" = "Balsam",
                         "BALSM" = "Balsam",
                         "BIRCH SP." = "Birch",
                         "YB." = "Birch",
                         "YB" = "Birch",
                         "yellow birch" = "Birch",
                         "PB" = "Birch",
                         "paper birch" = "Birch",
                         "BAL" = "Balsam",
                         "C "= "Cedar",
                         "C"= "Cedar",
                         "cedar"= "Cedar",
                         "H" ="Hemlock",
                         "H " = "Hemlock",
                         "WP" = "Pine",
                         "white pine?" = "Pine",
                         "IW" = "Ironwood",
                         "BASS" = "Basswood",
                         "WO?" = "Oak",
                         "WO" = "Oak",
                         "RO" = "Oak",
                         "O" = "Oak",
                         "BS" = "Spruce",
                         "SPRUCE SP." = "Spruce",
                         "RP" = "Pine",
                         "T" = "Tamarack",
                         "beech? Picture" = "Beech",
                         "BA" = "Ash",
                         "BA " = "Ash",
                         "BL" = "Black locust",
                         "ELM" = "Elm",
                         "UNKNWN" = "Unknown",
                         "UNKWN" = "Unknown",
                         "UNKNWN, APPLE?" = "Unknown"))
                                      
```

Get raw GEDI height

```{r}

#install.packages('rGEDI', repos = c('https://carlos-alberto-silva.r-universe.dev', 'https://cloud.r-project.org'))

# loading rGEDI package
library(rGEDI)



```

Get area boundaries

```{r}

# Approximate coordinates to cover all Apostle Islands
minlat <- 46.70  # slightly below Madeline Island
maxlat <- 47.05  # slightly above Sand Island
minlong <- -91.20  # west of Red Cliff
maxlong <- -90.50  # east of Devils Island


# minlat <- 40.45  # near Provo Canyon
# maxlat <- 41.80  # near Logan Canyon
# minlong <- -112.10  # west of Salt Lake City
# maxlong <- -110.60  # near Mirror Lake Highway and Evanston, WY

# Specifying the date range
daterange=c("2021-01-01","2021-01-31")

# # Get path to GEDI data
# gLevel2A <- gedifinder(
#   product = "GEDI02_A",
#   ul_lat = maxlat,
#   ul_lon = minlong,
#   lr_lat = minlat,
#   lr_lon = maxlong,
#   version = "002",
#   daterange = c("2021-01-01", "2021-12-31")
# )
# 
# # Just use the first few files
# gLevel2A_subset <- gLevel2A[1:5]  # for example, take just 5 files
```

```{r}

#gediDownload(filepath = gLevel2A, outdir = "Y:/rpienaar/Apostle Islands/Original Data Files/GEDI")

```

```{r}

# Read one file (update the filename to match one you downloaded)
#l2a_path <- "Y:/rpienaar/Apostle Islands/Original Data Files/GEDI/GEDI02_A_2021002075338_O11657_02_T08968_02_003_02_V002.h5"

# # List all GEDI L2A files in the folder
# files <- list.files("Y:/rpienaar/Apostle Islands/Original Data Files/GEDI", 
#                     pattern = "\\.h5$", full.names = TRUE)
# 
# # Function to read and clip each file
# read_clip_file <- function(file) {
#   gedi_l2a <- readLevel2A(file)
#   level2a <- getLevel2AM(gedi_l2a)
#   clipped <- clipLevel2AM(level2a, xmin = minlong, xmax = maxlong, ymin = minlat, ymax = maxlat)
#   as.data.frame(clipped)
# }
# 
# level2a_all <- bind_rows(lapply(files, read_clip_file))
# 
# 
# head(level2a_all[,c("beam","shot_number","elev_highestreturn","elev_lowestmode","rh100")])



```

```{r}
# Convert to data.frame
level2a_df <- as.data.frame(level2a_all)

level2a_df <- level2a_df[level2a_df$quality_flag == 1 &
                           level2a_df$degrade_flag < 1 & 
                           level2a_df$sensitivity < 1 &
                           level2a_df$rh100 >= 5,]

canopy_height <- level2a_df$rh100
ground_level <- level2a_df$rh0
lat <- level2a_df$lat_lowestmode
lon <- level2a_df$lon_lowestmode


# Combine into a data frame
canopy_df <- data.frame(lat, lon, rh100 = canopy_height, rh0 = ground_level)

# Preview
head(canopy_df)
filt <- canopy_df |> filter(rh0 >= 0)
plot(data = canopy_df,rh100~rh0)
hist(canopy_df$rh100)

#write.csv(canopy_df, "GEDI_RawFP.csv", row.names = FALSE)
```

```{r}

# Load required libraries
library(ggplot2)
library(maps)
library(leaflet)

# Plot with leaflet
# Add a new column for character labels
df$label_text <- paste("rh100:", df$rh100)

# Plot
leaflet(data = df) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius = 5,
    color = "blue",
    label = ~label_text,
    fillOpacity = 0.2
  )

```



```{r}

# Convert Height and DBH to numeric (in case they're read as characters)
df <- df %>%
  mutate(
    Height = as.numeric(Height),
    DBH = as.numeric(DBH)
  )

df <- df %>%
  filter(tolower(Vigor) != "snag") # removes snags

df <- df %>%
  mutate(Site = str_replace(Site, "^Oak", "OAK"),
         Site = str_replace(Site, "^Sand", "SAND")) # corrects names


head(df)

canopy_trees <- df |> filter(Height >= 15) # consider changing this

```






```{r}
# Compute canopy area proxy, this isnt canopy area though it is basal area of the stem/trunk
canopy_trees <- canopy_trees %>%
  mutate(canopy_area = (DBH^2))  

# Summarize by species at each site
species_canopy <- canopy_trees %>%
  group_by(Site, Species) %>%
  summarise(species_area = sum(canopy_area, na.rm = TRUE), .groups = "drop")

# Total canopy area per site
total_canopy <- species_canopy %>%
  group_by(Site) %>%
  summarise(total_area = sum(species_area), .groups = "drop")


#Add site coords back
site_coords <- df %>%
  select(Site, Lat, Long) %>%
  distinct()

canopy_proportion <- species_canopy %>%
  left_join(total_canopy, by = "Site") %>%
  mutate(proportion = species_area / total_area) %>%
  left_join(site_coords, by = "Site")  # <-- this keeps the lat/long


### Dominant species stands
stands <- filter(canopy_proportion, proportion >= 0.5)
head(stands)
stands$Species <- as.factor(stands$Species)
summary(stands$Species)
#write.csv(stands, "Stands.csv")

```


I want to determine the DBH cut off at which the tree is likely to be in the canopy

```{r}

ggplot(data= df, aes(x= DBH, y =Height, colour = Species))+
  geom_point()


# Fit the mixed-effects model
mod <- glm(Height ~ DBH * Species, data = df)

# Inspect the model
tab_model(mod)

# Define canopy height threshold
cutoff <- mean(canopy_trees$Height, na.rm = TRUE) + sd(canopy_trees$Height, na.rm = TRUE)



```



#NPS Ground Sampling

Load NPS data
```{r}

NPS <- read_csv("Y:/rpienaar/Apostle Islands/InputData/APIS_Veg_2021.csv")
NPS <- read_csv("Y:/rpienaar/Apostle Islands/InputData/TransectLocs.csv")




#NPSCanopy <- read.csv("CanopyHeightNPS.csv")

#colnames(NPSCanopy)[2] <- "Plot"


NPS <- NPS |> mutate(genus= stringr::word(Accepted_LatinName, 1))

genus_common <- tibble(
  genus = c("Acer", "Abies", "Betula", "Thuja", "Tsuga",
            "Quercus", "Fraxinus", "Populus", "Ostrya", "Picea", "Tilia", "Ulmus",
            "Prunus", "Sorbus", "Pinus", "unknown"),
  common_name = c("Maple", "Balsam", "Birch", "Cedar", "Hemlock",
                  "Oak", "Ash", "Aspen", "Ironwood", "Spruce", "Basswood", "Elm",
                  "Unknown", "Unknown", "Pine", "Unknown")
)



NPS <- NPS %>%
  left_join(genus_common, by = "genus")


NPS <- NPS |> select(Accepted_LatinName, Diameter_cm, Status, Plot, genus, common_name, Transect, SitePair, Northing, Easting )

```
Calculate heigth based on model from Monica

```{r}

#rename columns to match input data
colnames(NPS)[2] <- "DBH"
colnames(NPS)[6] <- "Species"

NPS$DBH <- as.numeric(NPS$DBH)
NPS$Species <- as.factor(NPS$Species)

NPS$Height <- predict(mod, newdata = NPS)


```


```{r}

NPS$Plot <- as.factor(NPS$Plot)
NPS <- NPS |> filter(Status == "Live")

NPScc <- NPS |> filter(Height >= 15)

# Create summary table for trees by transects
summary_table <- NPScc %>%
  group_by(Plot, Accepted_LatinName, Transect) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = Transect,
    values_from = Count,
    values_fill = 0  # fill missing combinations with 0
  ) %>%
  arrange(Plot, Accepted_LatinName)

NPScc$canopy <- NPScc$DBH^2

# Summarize by species at each site
NPSspecies_canopy <- NPScc %>%
  group_by(SitePair, Species) %>%
  summarise(species_area = sum(canopy, na.rm = TRUE), .groups = "drop")

# Total canopy area per site
NPStotal_canopy <- NPSspecies_canopy %>%
  group_by(SitePair) %>%
  summarise(total_area = sum(species_area), .groups = "drop")

#Add site coords back
NPSsite_coords <- NPS %>%
  select(SitePair, Northing, Easting) %>%
  distinct()

NPScanopy_proportion <- NPSspecies_canopy %>%
  left_join(NPStotal_canopy, by = "SitePair") %>%
  mutate(proportion = species_area / total_area) %>%
  left_join(NPSsite_coords, by = "SitePair")  # <-- this keeps the lat/long

NPScanopy_proportion$Northing <- as.numeric(NPScanopy_proportion$Northing)

```





Add common name

```{r}

NPScanopy_proportion |> filter(proportion > 0.5)

NPScanopy_proportion |> filter(Species == "Oak")


```


Now I have the 2 files of the 2 sources with the proportions of canopy cover, I should put them together to make one file

```{r}




NPScanopy_proportion <- na.omit(NPScanopy_proportion)
crs_utm <- 32615  
# Convert UTM to lat long
dfutmsf <- st_as_sf(NPScanopy_proportion, coords = c("Northing", "Easting"), crs = crs_utm)

dflatlong <- st_transform(dfutmsf, crs = 4326)

# Extract lat/long into columns
NPSprop <- dflatlong %>%
  mutate(Lat = st_coordinates(.)[,2],
          Long = st_coordinates(.)[,1]) %>%
  st_drop_geometry()


# Combine dfs
head(canopy_proportion)

head(NPSprop)

#rename cols
colnames(NPSprop)[1] <- "Site"
Combo <- rbind(canopy_proportion, NPSprop)

Combo$Species <- as.factor(Combo$Species)

summary(Combo$Species)

Combo <- Combo |> filter(!Species %in% c("Ash", "Balsam", "Basswood", "Beech", "Spruce"))

summary(Combo$Species)

head(Combo)
Combo$speciesID <- as.integer(factor(Combo$Species))

```


```{r}

ComboDom <- Combo |> filter(proportion >= 0.5) 

head(ComboDom)

Combo <- Combo |> group_by(Site, total_area, Lat, Long, Species) %>%
  summarise(prop_area = sum(species_area) / unique(total_area), .groups = "drop") %>%
  pivot_wider(
    names_from = Species,
    values_from = prop_area)
    
Combo[is.na(Combo)] <- 0

head(Combo)


#write.csv(ComboDom, "SpeciesClass.csv", row.names = FALSE)
#write.csv(Combo, "SpeciesProps.csv", row.names = FALSE)

```

Summary of data I have

```{r}
library(gt)

Combo |> pivot_longer(cols = Cedar:Hemlock, names_to = "Species", values_to = "Cover") %>%
  mutate(
    Category = case_when(
      Cover == 1 ~ "Pure stands",
      Cover > 0 & Cover < 1 ~ "Mixed stands",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Category != "Other") %>%
  count(Species, Category) %>%
  pivot_wider(names_from = Category, values_from = n, values_fill = 0) |>
  filter(`Pure stands` > 1) |>
  gt()

ComboDom |> 
  count(Species) %>%
  #pivot_wider(names_from = Category, values_from = n, values_fill = 0) |>
  filter(n > 1) |>
  gt()

```


Finish

```{r}


```

